---
title: "World regionalisation problem"
subtitle: "Experiments on trade flows (1967-2020)"
author: "Claude Grasland"
date: "2 Janvier 2024"
output :
  html_document:
    toc: true
    toc_depth: 3
bibliography: references.bib
---

## Objectifs


This working paper propose to discuss the theoretical problem of regionalisation of **a world (in abstract sense) ** through the empirical example of **The World (where we live)** described by trade flows over a long period of time and for different types of products.  

We will use for that purpose the CHELEM database produced by the CEPII which offers an exceptional coverage of trade flows over a period of 50 years from 1967 to present (2020). The most detailed version of this database describes the exchange between  94 x 94  territorial units (states or group of states) for 72 types of goods over a period of 54 years which means a 4-dimension object (hypercube) of size $94 \times 93 \times 72 \times 54 = 33988896$ cells. 

For our experiment, we will use a reduction of the database based on 12 territorial units described by 9 groups of goods for 5 periods of 10 years each. The hypercube used in our experiment will be therefore limited to a size of  $12 \times 11 \times 9 \times 5 = 5940$ cells. This can appear rather limited but - as we will demonstrate - the complexity of such an object is yet very high and it appears better to establish the theoretical foundation of the research of such an object before to adress larger databases where computational problem will grow exponentially. 

Our overarching question can now be formulated in the following way : 

- Let $W$ be **a world** divided in $1\dots i\dots  n$ territorial units.
- Let $F$ **a relation** defined on $W \times W$ which assign a value to each couple of units of the world (excluding only internal relations).
- Let $X$ **a typology of relations** in $1\dots k\dots  p$  types of relation using the same unit of mesurement.
- Let $T$ **a partition of time** in periods $1\dots t\dots  q$ where the relations are measured
- Let $H = F_{ijkt}$ the hypercube which measure of relation between territorial units $i$ and $j$ for the relation $k$ during time period $t$

- **Problem 1** : What are the partitions $P_i$ (for origins), $P_j$ (for destination), $P_k$ (for typology) and $P_t$ (for time period) that allows to reduce the size of the initial hypercube $H$ to a smaller one $H'$ without losing too much information. 
- **Problem 2** : can we identify homogeneous subparts of $H$ that are not necessarily based on orthogonal divisions of the hypercube. 
- **Problem 3**: can we identify trajectories of regionalisation $P_i(t)$,  $P_j(j)$  or trajectories of typology $P_k(t)$ that descibe the evolution of optimal partitions through time . 




```{r, echo=F, warning=F, message=F, error=F}
library(sf,quietly = T)
library(mapsf)
library(dplyr, quietly=T)
library(data.table, quietly = T)
library(ggplot2)
library(ggpubr)
library(plotly)
library(RColorBrewer)
library(visNetwork) 
library(knitr)
theme_set(theme_pubr())
knitr::opts_chunk$set(echo = FALSE,comment = "",warning = FALSE,error = FALSE,message = FALSE)


```

## I. THE EXPERIMENTAL DATABASE



```{r load}
dt<-readRDS("data/MYCHELEM_V2.RDS")
dt$t<-as.numeric(dt$t)
```


### The 94 terrtorial units of CHELEM

The original version of the CHELEM database is made of 94 territorial units. A majority of this territorial units correspond to states but some of them are made of aggregates of states for which it was difficult to separate trade flows or to collect them. The map below indicates what are the territorial units that do not fit with international division of the world in states. 

```{r}

chelem<-readRDS("data/World_Chelem.RDS") 

ns<-c("MLT","BLX","MOA","AFA","AFZ","SAF","AMA","AOZ","CAL","AOA","CAU","CEA","GOA")
agr<-chelem %>% filter(code %in% ns)


mypal<-rainbow(20)
mf_theme("agolalight")
mf_map(chelem, type="base", col="lightyellow")
mf_map(agr,
       type = "typo",
       var="name",
       leg_pos = "left",
       leg_title = "Groupes",
       add=T,
       pal=mypal
       )

mf_layout(title = "Agrégats de pays dans la base CHELEM",
          credits = "Source : CEPII, Base CHELEM",
          frame = T,
          arrow = F
          )

```

The aggregates of states are generally based on groups of small states (like in central America or Oceania) but it can also be the case for larger goups of states playing an important role in trade like in the case of the aggregate between Irag, Iran and Koweit. The aggregation is also very large in the case of subsaharan Africa where only few states are identified and the other mixed in large area, not necessarily contiguous. At the same time, Europe is fully disaggregated in isolated states, except in the case of Malta and Cyprus, which will have for consequence an increase of trade flows in this part of the world. if USA was divided in 51 federal states and China or India in provinces or states, it would necessarily increase their part of exchanges. 

We are therefore facing here a difficult question of Modifiable Area Unit Problem (MAUP) which can not be easily solved without deciding immediately to aggregate the data in larger units, more homogeneous,  where internal flows will be systematically removed. This will produce of course a strong reduction of the initial information but make possible to have a better analysis of the relation between the new territorial units. 


### The 4 x3 =  12 basic territorial units


On the basis of expert advices, we have chosen 12 basic territorial units which are in fact associated to a first division of the world in 4 regions, each of them divided in 3 subregions. 

```{r}

gui2<-readRDS("data/Chelem_Reg_2.RDS") %>% filter(GUI2 !="G01")
gui2$code<-paste(gui2$GUI2,gui2$GUI2_Name)


mypal<-c(rev(brewer.pal(5,"Blues")[1:3]),
         rev(brewer.pal(5, "Reds")[1:3]),
        rev(brewer.pal(5,"YlOrBr")[1:3]),
         rev(brewer.pal(5,"Greens")[1:3])

         )
#mypal<-rainbow(12)
mf_theme("agolalight")
mf_map(gui2,
       type = "typo",
       var="code",
       leg_pos = "left",
       leg_title = "Régions",
       pal=mypal,
       )

mf_layout(title = "The 12 basic territorial units",
          credits = "Source : C. Grasland, 2024",
          frame = T,
          arrow = F
          )

```

The autors of this partition of the world suggest that the world economy has been (at least during a period of time) or could have been (whishfull thinking ?) be organized around three integrated "vertical macroregions" and one residual part of the word less integrated and submit to variable influence of the three vertical regions : 

- **G1 : Europe-Mediterranea-Africa** :  Clearly inherited from the history, this vertical region is based on various type of proximities including geographical distance, common sea (Mare Nostrum), common language, colonial legacy ... But what has been the destiny of these links over the last 50 years following the independancy of states from Africa ?
- **G2 : Americas** : Since the 19th century, *"the Monroe Doctrine is a United States foreign policy position that opposes European colonialism in the Western Hemisphere. It holds that any intervention in the political affairs of the Americas by foreign powers is a potentially hostile act against the United States"* (Wikipedia). This doctrine has been related to lot of conflict between the different parts of Americas but also associated to the building of various forms of cooperation like NAFTA (1994), MERCOSUR (1991), etc... In any case, the geographical proximity was clearly here in favor of a potential integration. But the reduction of transport cost in the 1980's has modified the role played by these factor in favor of trans-Pacific relationships. So, what is the situation of America's integration over our 50 years period of interest ?
- **G3 : Asia-Pacifica** : The economic integration of this part of the world is a long and complex process initially boosted by Japan and Korea, further by China and associated to a continuous process or development of free trade areas like ASEAN. This potential macro-region has been at the same time the pivot of global economic integration of the world, firstly with trans-pacific relation until 1990 and further with the rest of the world with the growing influence of China after this state joined the WTO in  2001. So, is it still a macroregion or the economic core of contemporary world ?

- **G4 : Rest of the World** : We can not speak here from an integrated economic region but rather as a group of states that (1) benefit from ressources of interest forthe rest of the world (e.g. oil and gas from the Gulf, mineral products from Russia, ...) and/or (2) develop a strategy of diversification of their exchange at world scale and refuse to be dependent from too powerful partners (e.g. strategy of India, Russia or Saudi Arabia). The question here is to what extent this part of the world remained "neutral" as compare to the three other ones or has been succesfully associated to the different other regions according to variable geometries.

All this remarks are hypothesis that suggest a possible way to cluster the 12 territorial units in 3 or four groups. But our aim is not here to validate the partition $(G_1,G_2,G_3,G_4)$ but rather to use it at starting point for the discovery of alternative geometries changing throug time or presenting variable configurations according to the type of products considered. 


### 9 groups of goods

The authors of the database CHELEM as made incredible efforts to maintain an homogeneous categorisation of goods in 72 types of producst over a period of 50 years. Considering the changes of the world economy and the evolution of the nomenclature used by trade organization, it is a genuine miracle to have done such a work. We adopt here a simplified version of the CHELEM typology in only 9 groups of products that reflect the distribution of value chains as well as the international division of labor  (@grasland2010, @grataloup2014)

- (1) ENE : Energy
- (2) MIN : Minerals, Intermediate goods
- (3) AGR : Agriculture, Food
- (4) TEX : Textile, Clothing
- (5) ELE : Electronic
- (6) EQU : Equipment, Machines
- (7) TRA : Transport
- (8) CHE : Chemical products
- (9) MIS : Others



```{r}
Fkt<-dt[k!="TOT",.(Fkt=sum(Fijkt)),.(k,t)]
Ft<-dt[k!="TOT",.(Ft=sum(Fijkt)),.(t)]
tab<-left_join(Fkt,Ft) %>% mutate(pct=100*Fkt/Ft)
g<-ggplot(tab) + aes(x=t, y=pct,fill=k) +
               geom_bar(stat="identity") +
       scale_x_continuous("Année") +
       scale_y_continuous("% flux mondiaux") +
       ggtitle("Typology of trade goods in 9 groups",subtitle = "Source : CEPII, Base CHELEM")
ggplotly(g)
```

- **Commentaires** : A statistical interest of this typology (out of the fact that it is relevant in terms of divsion of labor) is that the groups are relatively equlibrated in size. They offer interesting trends of variation of their respective shares that can increase (electronic, chemical), decrease (agriculture) or present chaotic evolution related to the variations of price (energy)

### The hypercube with 5940 cells

```{r, eval=FALSE}
# This program build the hypercube

dt2<-dt %>% mutate(t=as.numeric(t)) %>%filter(t>1970)
dt2$t<-cut(dt2$t,breaks=c(1970, 1980,1990, 2000, 2010, 2020 ))
table(dt2$t)
levels(dt2$t)<-c("1971-80","1981-90","1991-00","2001-10","2011-20")
hc<-dt2[, .(Fijkt = sum(Fijkt)), .(Gi=reg1i, i=reg2i, Gj=reg1j, j=reg2j, k,t)]
hc<-hc %>% filter(i !=j, k!="TOT")
hc2<-hc %>% group_by(t) %>% mutate(Fijkt=round(Fijkt*1000000/sum(Fijkt),0)) %>% select(i,j,k,t,Fijkt)
hc3<-hc2 %>% select(i,j,k,t,Fijkt)
names(hc3)<-c("j","i","k","t","Fjikt")
hc4<-left_join(hc2,hc3) %>% arrange(i,j,k,t) %>% mutate(Vijkt=(Fijkt+Fjikt)/2, Bijkt=Fijkt-Fjikt)
saveRDS(hc4,"data/MYCHELEM_Hypercube.RDS")
```

On the basis of previous rules we have built the expected hypercube with 5940 cells. The flows has been normalized to an arbitrary total sum of 1000000 for each period of ten years and the values has been round with zero decimal. We have introduced for each couple of region the flows in both direction $F_{ijkt}$ and $F_{jikt}$ in order to be able to compute easily the symetric part of exchange called volume and the asymetric part called balance :

- Volume : $V_{ijkt} = \frac{(F_{ijkt}+F_{jikt})}{2}$
- Balance : $B_{ijkt} = F_{ijkt}-F_{jikt}$


```{r}
hc<-readRDS("data/MYCHELEM_Hypercube.RDS")
kable(head(hc,10))
```
 

